---
header-includes:
   - \usepackage{booktabs,float,longtable }
   - \hypersetup{breaklinks=true}
   - \usepackage{breakurl}
output: 
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
title: "A Pandoc Markdown Article Starter and Template"
thanks: "Replication files are available on the author's Github account..."
author:
- name: Steven V. Miller
  affiliation: Clemson University
- name: Mary Margaret Albright
  affiliation: Pendelton State University
- name: Rembrandt Q. Einstein
  affiliation: Springfield University
abstract: "This document provides an introduction to R Markdown, argues for its..."
keywords: "pandoc, r markdown, knitr"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt


---


```{r setup, include=FALSE}

rm(list=ls(all=TRUE)) #start with empty workspace

startTime <- Sys.time() # get the time

knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, warning = FALSE, include = FALSE)

library(RODBC) # package to import data from Excel or Access
library(tidyverse) # package to clean up/manipulate data
library(sf) # to work with maps
library(sp)
library(hrbrthemes)

library(httr); set_config(config(ssl_verifypeer = 0L)); devtools::source_url("https://raw.githubusercontent.com/coquito77/R_functions/master/abbFileName.R")

setwd("~/GitHub/2016ElectionsCntyLevlChrpIn/") # set the working directory
list.files() # see whats there


```


```{r getdata, eval = FALSE}

# get votes data

voteData <- read.csv("https://raw.githubusercontent.com/bcdunbar/datasets/master/votes.csv") %>% 
  mutate(combined_fips = as.factor(as.character(combined_fips)),
         nchartest = nchar(as.character(combined_fips)),
         combined_fips = ifelse(nchartest == 4, paste0("0",combined_fips),as.character(combined_fips)),
         combined_fips = as.factor(as.character(combined_fips))) %>% 
  select(votes_dem_2016,votes_gop_2016,total_votes_2016,Clinton,
         Trump,diff_2016,per_point_diff_2016,state_abbr, 
         county_name, combined_fips, combined_fips)

# get county data
 download.file('http://www2.census.gov/geo/tiger/GENZ2016/shp/cb_2016_us_county_500k.zip', 'county.zip',
              quiet = TRUE)

temp <- unzip('county.zip')

projections <- st_proj_info(type = "proj")

cb_2016_us_county_500k <- read_sf("./cb_2016_us_county_500k.shp", quiet = TRUE, stringsAsFactors = FALSE) 

st_crs(cb_2016_us_county_500k)

save.image(file = 'data.RData')

```

```{r processData, eval = TRUE}
# setwd("~/GitHub/2016ElectionsCntyLevlChrpIn/") ; list.files() # see whats there

load('data.RData')

# need to get data for missing county 
# https://en.wikipedia.org/wiki/Oglala_Lakota_County,_South_Dakota

myurl <- "https://uselectionatlas.org/RESULTS/statesub.php?year=2016&fips=46113&f=1&off=0&elect=0"

OglalaLakota <- xml2::read_html(GET(myurl, user_agent("myagent"))) %>%
  rvest::html_table(fill = TRUE,header = NA) %>%
  .[[3]]  %>% as.tibble %>% select(PoliticalParty = X4, PopularVote =  X5) %>% 
  slice( 3:6) %>%   mutate(county_name = "Oglala Lakota",
                           combined_fips =  "46102",
                           state_abbr = "SD") %>% 
  spread(PoliticalParty,PopularVote) %>% 
  select(county_name, combined_fips, state_abbr, votes_dem_2016 =  Democratic,
         votes_gop_2016 =  Republican, Libertarian, Constitution) %>% 
  mutate_at(vars(votes_dem_2016:Constitution),
            funs(as.numeric(gsub(",","",.)))) %>% 
  mutate(total_votes_2016 = votes_gop_2016 + votes_dem_2016 +Libertarian+ Constitution,
         Clinton = votes_dem_2016/total_votes_2016,
         Trump = votes_gop_2016/total_votes_2016,
         diff_2016 = votes_gop_2016- votes_dem_2016,
         per_point_diff_2016 = Clinton- Trump ) %>% 
  select(-c( Libertarian, Constitution))

voteData <- rbind(voteData, OglalaLakota) %>%
  mutate(Won = ifelse(Clinton > Trump,"Clinton","Trump"))

countyVotes <- cb_2016_us_county_500k %>% 
  st_transform(., 2229) %>% 
  mutate(combined_fips = paste0(STATEFP, COUNTYFP),
         nchartest = nchar(as.character(combined_fips))) %>% as_tibble() %>% 
  inner_join(voteData, by = "combined_fips") %>% 
  select(STATEFP, COUNTYFP,COUNTYNS,AFFGEOID,GEOID,NAME,
         votes_dem_2016,votes_gop_2016,total_votes_2016,Clinton,
         Trump,diff_2016,per_point_diff_2016,state_abbr, 
         county_name, geometry, Won) %>% 
  mutate_if(is.character, as.factor)

st_crs(countyVotes)

try(st_write(countyVotes, "countyVotes.shp"))

# disolve map by state

stateVotes <- countyVotes %>% 
  group_by(state_abbr) %>% 
  summarize(TotVotes_dem_2016 = sum(votes_dem_2016, na.rm = TRUE),
            TotVotes_gop_2016 = sum(votes_gop_2016, na.rm = TRUE),
            Tottotal_votes_2016 = sum(total_votes_2016, na.rm = TRUE),
            TotClinton = TotVotes_dem_2016/Tottotal_votes_2016,
            TotTrump = TotVotes_gop_2016/Tottotal_votes_2016) %>% 
  ungroup() %>%
  mutate(Won = ifelse(TotClinton > TotTrump,"Clinton","Trump"))

st_crs(stateVotes)

try(st_write(stateVotes, "stateVotes.shp")) 

```

```{r loadMaps}


library(extrafont)
font_import(pattern = "Roboto", prompt=FALSE)
loadfonts(device = "win", quiet = TRUE)
import_roboto_condensed()

update_geom_font_defaults(family=font_rc_light)

# state
# 
# state map

stateMapbreaks <- fivenum(stateVotes$TotClinton, na.rm = TRUE) %>% round(.,2)

stateVotesMap <- as(stateVotes, 'Spatial')

proj4string(stateVotesMap) <- CRS("+init=epsg:2229")

stateFrtMap <- broom::tidy(stateVotesMap, region = "state_abbr")
  
stateFrtMap <- dplyr::left_join(stateFrtMap,
                                       stateVotesMap@data,
                                       by = c("id" = "state_abbr"))

# state cartogram

stateVoteCartogram <- read_sf("./stateCartogram.shp", quiet = TRUE, stringsAsFactors = FALSE) 

stateVoteCartogram <- as(stateVoteCartogram, 'Spatial')

proj4string(stateVoteCartogram) <- CRS("+init=epsg:2229")

stateFrtCartogram <- broom::tidy(stateVoteCartogram, region = "stt_b")
  
stateFrtCartogram <- dplyr::left_join(stateFrtCartogram,
                                       stateVoteCartogram@data,
                                       by = c("id" = "stt_b"))

# county map
#
#  map
countyMapbreaks <- fivenum(countyVotes$Clinton, na.rm = TRUE) %>% round(.,2)

countyVotesMap <- as(countyVotes, 'Spatial')

proj4string(countyVotesMap) <- CRS("+init=epsg:2229")

countyFrtMap <- broom::tidy(countyVotesMap, region = "AFFGEOID")
  
countyFrtMap <- dplyr::left_join(countyFrtMap,
                                       countyVotesMap@data,
                                       by = c("id" = "AFFGEOID"))


# county cartogram

countyVoteCartogram <- read_sf("./Cartogram.shp", quiet = TRUE, stringsAsFactors = FALSE) 

countyVoteCartogram <- as(countyVoteCartogram, 'Spatial')

proj4string(countyVoteCartogram) <- CRS("+init=epsg:2229")

countyFrtCartogram <- broom::tidy(countyVoteCartogram, region = "AFFGEOI")
  
countyFrtCartogram <- dplyr::left_join(countyFrtCartogram,
                                       countyVoteCartogram@data,
                                      by = c("id" = "AFFGEOI"))


# to do
# create ggplot theme options



# create maps
# map state discrete 

mapStateDisc <- ggplot() +
  geom_polygon(data = stateFrtMap,
               aes(long, lat, group = group, fill = Won),
               # fill="grey40",
               colour = "grey90", 
               #alpha = .7,
               size = .015) +
  #theme_ipsum_rc(plot_title_size = 15, subtitle_size = 10, caption_size = 5) +
  theme(#aspect.ratio = 1,
                 panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(),
                 panel.background = element_rect(#colour = "grey90",
                   size = 1,
                   fill = "transparent",colour = NA),
                 strip.text.x = element_text(size = 5,
                                             colour = "black",
                                             angle = 00),
                 legend.position = "bottom", # c(.55, .10),
                 #plot.margin = unit(c(0,0,-.5,-.5), "cm"),
                 legend.key.width = unit(1, "cm"),
                 axis.title.x = element_blank(),
                 axis.text.x = element_blank(),
                 axis.title.y = element_blank(),
                 axis.text.y = element_blank(),
                 #legend.key.width = unit(3,"line"),
                 plot.margin = unit(c(3,3,3,3),"mm")) +
  labs(#x = "Year",
    #y = "yr""
    title = "test",
    subtitle = paste("test"), 
    caption  = "Source: test"
  )

# map state continous

mapStateCont <- ggplot() +
    geom_polygon(data = stateFrtMap,
                 aes(long, lat, group = group, fill = TotClinton),
                 # fill="grey40",
                 colour = "grey90",
                 #alpha = .7,
                 size = .015) +
  scale_fill_gradient2(name = "Percent",
                       breaks = stateMapbreaks, labels =  stateMapbreaks)  +
  #theme_ipsum_rc(plot_title_size = 15, subtitle_size = 10, caption_size = 5) +
  theme(#aspect.ratio = 1,
                 panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(),
                 panel.background = element_rect(#colour = "grey90",
                   size = 1,
                   fill = "transparent",colour = NA),
                 strip.text.x = element_text(size = 5,
                                             colour = "black",
                                             angle = 00),
                 legend.position = "bottom", # c(.55, .10),
                 #plot.margin = unit(c(0,0,-.5,-.5), "cm"),
                 legend.key.width = unit(1, "cm"),
                 axis.title.x = element_blank(),
                 axis.text.x = element_blank(),
                 axis.title.y = element_blank(),
                 axis.text.y = element_blank(),
                 #legend.key.width = unit(3,"line"),
                 plot.margin = unit(c(3,3,3,3),"mm")) +
  labs(#x = "Year",
    #y = "yr""
    title = "test",
    subtitle = paste("test"), 
    caption  = "Source: test"
  )

# cartogram state discrete 

crtStateDisc <- ggplot() +
    geom_polygon(data = stateFrtCartogram,
                 aes(long, lat, group = group, fill = Won),
                 # fill="grey40",
                 colour = "grey90",
                 #alpha = .7,
                 size = .015)  +
  #theme_ipsum_rc(plot_title_size = 15, subtitle_size = 10, caption_size = 5) +
  theme(#aspect.ratio = 1,
                 panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(),
                 panel.background = element_rect(#colour = "grey90",
                   size = 1,
                   fill = "transparent",colour = NA),
                 strip.text.x = element_text(size = 5,
                                             colour = "black",
                                             angle = 00),
                 legend.position = "bottom", # c(.55, .10),
                 #plot.margin = unit(c(0,0,-.5,-.5), "cm"),
                 legend.key.width = unit(1, "cm"),
                 axis.title.x = element_blank(),
                 axis.text.x = element_blank(),
                 axis.title.y = element_blank(),
                 axis.text.y = element_blank(),
                 #legend.key.width = unit(3,"line"),
                 plot.margin = unit(c(3,3,3,3),"mm")) +
  labs(#x = "Year",
    #y = "yr""
    title = "test",
    subtitle = paste("test"), 
    caption  = "Source: test"
  )

# cartogram state continous

crtStateCont <- ggplot() +
  geom_polygon(data = stateFrtCartogram,
               aes(long, lat, group = group, fill = TtCln),
               # fill="grey40",
               colour = "grey90",
               #alpha = .7,
               size = .015) +
  scale_fill_gradient2(name = "Percent",
                       breaks = stateMapbreaks, labels =  stateMapbreaks) +
  #theme_ipsum_rc(plot_title_size = 15, subtitle_size = 10, caption_size = 5) +
  theme(#aspect.ratio = 1,
                 panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(),
                 panel.background = element_rect(#colour = "grey90",
                   size = 1,
                   fill = "transparent",colour = NA),
                 strip.text.x = element_text(size = 5,
                                             colour = "black",
                                             angle = 00),
                 legend.position = "bottom", # c(.55, .10),
                 #plot.margin = unit(c(0,0,-.5,-.5), "cm"),
                 legend.key.width = unit(1, "cm"),
                 axis.title.x = element_blank(),
                 axis.text.x = element_blank(),
                 axis.title.y = element_blank(),
                 axis.text.y = element_blank(),
                 #legend.key.width = unit(3,"line"),
                 plot.margin = unit(c(3,3,3,3),"mm")) +
  labs(#x = "Year",
    #y = "yr""
    title = "test",
    subtitle = paste("test"), 
    caption  = "Source: test"
  )

# map county discrete 
mapCountyDisc <- ggplot() +
  geom_polygon(data = countyFrtMap,
               aes(long, lat, group = group, fill = Won),
               # fill="grey40",
               colour = "grey90",
               #alpha = .7,
               size = .015)+
  #theme_ipsum_rc(plot_title_size = 15, subtitle_size = 10, caption_size = 5) +
  theme(#aspect.ratio = 1,
                 panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(),
                 panel.background = element_rect(#colour = "grey90",
                   size = 1,
                   fill = "transparent",colour = NA),
                 strip.text.x = element_text(size = 5,
                                             colour = "black",
                                             angle = 00),
                 legend.position = "bottom", # c(.55, .10),
                 #plot.margin = unit(c(0,0,-.5,-.5), "cm"),
                 legend.key.width = unit(1, "cm"),
                 axis.title.x = element_blank(),
                 axis.text.x = element_blank(),
                 axis.title.y = element_blank(),
                 axis.text.y = element_blank(),
                 #legend.key.width = unit(3,"line"),
                 plot.margin = unit(c(3,3,3,3),"mm")) +
  labs(#x = "Year",
    #y = "yr""
    title = "test",
    subtitle = paste("test"), 
    caption  = "Source: test"
  )

# map county continous

mapCountyCont <- ggplot() +
  geom_polygon(data = countyFrtMap,
               aes(long, lat, group = group, fill = Clinton),
               # fill="grey40",
               colour = "grey90",
               #alpha = .7,
               size = .015) +
  # coord_equal(xlim = c(6668481, 6275488), 
  #             ylim = c( 2122084,  16.5e+05))  +
  scale_fill_gradient2(name = "Percent",
                       breaks = countyMapbreaks, labels =  countyMapbreaks) +
  #theme_ipsum_rc(plot_title_size = 15, subtitle_size = 10, caption_size = 5) +
  theme(#aspect.ratio = 1,
                 panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(),
                 panel.background = element_rect(#colour = "grey90",
                   size = 1,
                   fill = "transparent",colour = NA),
                 strip.text.x = element_text(size = 5,
                                             colour = "black",
                                             angle = 00),
                 legend.position = "bottom", # c(.55, .10),
                 #plot.margin = unit(c(0,0,-.5,-.5), "cm"),
                 legend.key.width = unit(1, "cm"),
                 axis.title.x = element_blank(),
                 axis.text.x = element_blank(),
                 axis.title.y = element_blank(),
                 axis.text.y = element_blank(),
                 #legend.key.width = unit(3,"line"),
                 plot.margin = unit(c(3,3,3,3),"mm")) +
  labs(#x = "Year",
    #y = "yr""
    title = "test",
    subtitle = paste("test"), 
    caption  = "Source: test"
  )


# cartogram county discrete

mapCountCont <- ggplot() +
  geom_polygon(data = countyFrtCartogram,
               aes(long, lat, group = group, fill = Won),
               # fill="grey40",
               colour = "grey90",
               #alpha = .7,
               size = .015)+
  #theme_ipsum_rc(plot_title_size = 15, subtitle_size = 10, caption_size = 5) +
  theme(#aspect.ratio = 1,
                 panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(),
                 panel.background = element_rect(#colour = "grey90",
                   size = 1,
                   fill = "transparent",colour = NA),
                 strip.text.x = element_text(size = 5,
                                             colour = "black",
                                             angle = 00),
                 legend.position = "bottom", # c(.55, .10),
                 #plot.margin = unit(c(0,0,-.5,-.5), "cm"),
                 legend.key.width = unit(1, "cm"),
                 axis.title.x = element_blank(),
                 axis.text.x = element_blank(),
                 axis.title.y = element_blank(),
                 axis.text.y = element_blank(),
                 #legend.key.width = unit(3,"line"),
                 plot.margin = unit(c(3,3,3,3),"mm")) +
  labs(#x = "Year",
    #y = "yr""
    title = "test",
    subtitle = paste("test"), 
    caption  = "Source: test"
  )

# cartogram county continous

crtCountCont <- ggplot() +
  geom_polygon(data = countyFrtCartogram,
               aes(long, lat, group = group, fill = Clinton),
               # fill="grey40",
               colour = "grey90",
               #alpha = .7,
               size = .015) +
  # coord_equal(xlim = c(6668481, 6275488), 
  #             ylim = c( 2122084,  16.5e+05))  +
  scale_fill_gradient2(name = "Percent",
                       breaks = countyMapbreaks, labels =  countyMapbreaks)+
  #theme_ipsum_rc(plot_title_size = 15, subtitle_size = 10, caption_size = 5) +
  theme(#aspect.ratio = 1,
                 panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(),
                 panel.background = element_rect(#colour = "grey90",
                   size = 1,
                   fill = "transparent",colour = NA),
                 strip.text.x = element_text(size = 5,
                                             colour = "black",
                                             angle = 00),
                 legend.position = "bottom", # c(.55, .10),
                 #plot.margin = unit(c(0,0,-.5,-.5), "cm"),
                 legend.key.width = unit(1, "cm"),
                 axis.title.x = element_blank(),
                 axis.text.x = element_blank(),
                 axis.title.y = element_blank(),
                 axis.text.y = element_blank(),
                 #legend.key.width = unit(3,"line"),
                 plot.margin = unit(c(3,3,3,3),"mm")) +
  labs(#x = "Year",
    #y = "yr""
    title = "test",
    subtitle = paste("test"), 
    caption  = "Source: test"
  )


```


```{r plotPresidentsHeights, include =  FALSE, eval = FALSE, message=FALSE, cache = FALSE}


library(grid)
library(gridExtra)

# grid.newpage()
# pushViewport(viewport(layout = grid.layout(2, 1))) # row and col
# 
# print(mapStateDisc, vp = vplayout(1, 1))
# print(mapStateCont, vp = vplayout(2, 1))


pushViewport(viewport(layout = grid.layout(nrow = 2, ncol = 1, heights = unit(c(4, 4), "null")))) 
print(mapStateDisc, vp = viewport(layout.pos.row = 1, layout.pos.col = 1))
print(mapStateCont, vp = viewport(layout.pos.row = 2, layout.pos.col = 1))



```


```{r plotcombo, include =  TRUE, eval = TRUE, message = FALSE, cache = TRUE, fig.width=6, fig.height=9}

library(extrafont)
font_import(pattern = "Roboto", prompt=FALSE)
loadfonts(device = "win", quiet = TRUE)
import_roboto_condensed()

update_geom_font_defaults(family=font_rc_light)

library(cowplot)

plot_grid(mapStateDisc,
          mapStateCont,
          crtStateDisc,
          crtStateCont,
          labels = c("mapStateDisc","mapStateCont",
                     "crtStateDisc","crtStateCont"),
          label_size = 10,
          ncol = 2,
          rel_widths = .5,
          rel_heights = 1)


```



```{r plotcombo2, include =  TRUE, eval = TRUE, message = FALSE, cache = TRUE, fig.width=6, fig.height=9}

library(extrafont)
font_import(pattern = "Roboto", prompt=FALSE)
loadfonts(device = "win", quiet = TRUE)
import_roboto_condensed()

update_geom_font_defaults(family=font_rc_light)

library(cowplot)

plot_grid(mapCountyDisc,
          mapCountyCont,
          mapCountCont,
          crtCountCont,
          labels = c("mapCountyDisc","mapCountyCont",
                     "mapCountCont","crtCountCont"),
          label_size = 10,
          ncol = 2,
          rel_widths = .5,
          rel_heights = 1)


```

```{r sample, echo=FALSE, results='asis', include= TRUE}

# from here
# https://stackoverflow.com/questions/13675380/inset-footnote-into-header-with-xtable-and-tabular-environment

require(xtable)
x <- matrix(rnorm(60), ncol = 10)
x.big <- xtable(x,label='tabbig', caption='Example of longtable')
names(x.big) <- LETTERS[1:10]
names(x.big)[9] <- paste('I','footnote')    # I put the tag on I letter 
print(x.big,tabular.environment='longtable',floating=FALSE,
      sanitize.text.function=function(str)gsub("footnote","\\footnote{my tricky footnote !!}",str,fixed=TRUE))
      
```

I'm writing something here to test \footnote{footnotes working fine}
several features.


\begin{thebibliography}{9}

\bibitem{lamport94}
  Leslie Lamport,
  \textit{\LaTeX: a document preparation system},
  Addison Wesley, Massachusetts,
  2nd edition,
  1994.

\bibitem{latexcompanion} 
Michel Goossens, Frank Mittelbach, and Alexander Samarin. 
\textit{The \LaTeX\ Companion}. 
Addison-Wesley, Reading, Massachusetts, 1993.
 
\bibitem{einstein} 
Albert Einstein. 
\textit{Zur Elektrodynamik bewegter K{\"o}rper}. (German) 
[\textit{On the electrodynamics of moving bodies}]. 
Annalen der Physik, 322(10):891–921, 1905.
 
\bibitem{knuthwebsite} 
Knuth: Computers and Typesetting,
\url{http://www.rivcoems.org/Portals/0/Documents/DOCUMENTS/CONTRACTS/AMR/Current%20AMR%20Contract%20%202015.pdf}


\end{thebibliography}
